---
title: "R Basics & The Weighted Dice"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: "A first hour in R: arithmetic, objects, vectors, functions, and building a pair of loaded dice."
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.timelimit = 30)
set.seed(NULL)
```

## Welcome

Today you are going to build a pair of dice in R --- and then quietly weight them
so they roll 12 more often than they should.

Along the way you will meet almost everything you need to start using R:

- arithmetic and **objects**
- **vectors** and element-wise math
- **functions** and their **arguments**
- writing a function of your own
- repeating a simulation thousands of times

**How this works.** Every grey box below is a live R console. Type into it and
click **Run Code**. Nothing you do here can break anything --- if a box gets into
a mess, click **Start Over** at the top of the section and try again.

Boxes with a **Hint** button will nudge you; **Solution** shows the answer. Try
it yourself before you peek.

## 1. R is a calculator

At its simplest, R evaluates what you type and prints the answer.

Run this box as-is first:

```{r calc1, exercise=TRUE}
1 + 1
```

Now **change it**. In the box below, work out how many minutes are in a week.

```{r calc2, exercise=TRUE}

```

```{r calc2-hint}
# minutes in an hour, times hours in a day, times days in a week
60 * 24
```

```{r calc2-solution}
60 * 24 * 7
```

R has the operators you would expect: `+  -  *  /  ^`. It also has some you may
not: `%%` gives the remainder after division.

```{r calc3, exercise=TRUE}
17 %% 5
2^10
```

### An incomplete command

Type `5 -` into the box and run it.

```{r calc4, exercise=TRUE}
5 -
```

R answers with a `+` prompt instead of a result. That is R saying *"I'm still
waiting --- you haven't finished."* In your own console, press **Escape** to
cancel and start again. This is the single most common thing that confuses
people in their first week, and now it will not confuse you.

```{r quiz-prompt}
quiz(caption = "Quick check",
  question("You type a command and R responds with a `+` prompt instead of an answer. What happened?",
    answer("The command is incomplete --- R is waiting for the rest of it.", correct = TRUE),
    answer("R is adding something."),
    answer("An error occurred and R has crashed."),
    answer("R is asking you to confirm the command."),
    allow_retry = TRUE,
    correct = "Right. Press Escape to cancel, then retype the full command.",
    incorrect = "Not quite --- the `+` prompt means R thinks your command isn't finished yet."
  )
)
```

## 2. Objects

You can store a result in an **object** with the arrow `<-`. R's keyboard
shortcut for the arrow is **Alt + -** (Windows) or **Option + -** (Mac).

```{r obj1, exercise=TRUE}
a <- 1
a
```

Notice that assigning does *not* print anything. To see an object, type its
name.

Once an object exists, you can use it anywhere you would use its value:

```{r obj2, exercise=TRUE}
a <- 1
a + 2
a <- 100
a + 2
```

Your turn. Create an object called `die` that holds the numbers 1 through 6.
The colon operator makes a sequence: `1:6`.

```{r obj3, exercise=TRUE}

```

```{r obj3-hint}
# use the arrow, then print the object to check
die <- ...
```

```{r obj3-solution}
die <- 1:6
die
```

R is **case sensitive**: `die` and `Die` are two different objects. Names cannot
begin with a number, and cannot contain spaces or most symbols.

```{r quiz-names}
quiz(caption = "Which of these are legal object names?",
  question("Select every name R will accept.",
    answer("`my_die`", correct = TRUE),
    answer("`die2`", correct = TRUE),
    answer("`2die`", message = "Names cannot start with a number."),
    answer("`my die`", message = "Names cannot contain spaces."),
    answer("`.die`", correct = TRUE),
    allow_retry = TRUE
  )
)
```

## 3. Vectors do maths all at once

`die` is not a single number --- it is a **vector** of six numbers. When you do
arithmetic on a vector, R does it to *every element*.

Predict what this prints before you run it:

```{r vec1, exercise=TRUE}
die <- 1:6
die - 1
die / 2
die * die
```

That last one may have surprised you. `die * die` did **not** do matrix
multiplication --- it multiplied each element by the element in the same
position: 1\*1, 2\*2, 3\*3, and so on. This is called **element-wise** execution
and it is everywhere in R.

### Recycling

What if the two vectors are different lengths? R **recycles** the shorter one
until it is long enough.

```{r vec2, exercise=TRUE}
die <- 1:6
die + 1:2
```

`1:2` became `1 2 1 2 1 2`. Useful, and occasionally a trap --- if the lengths
don't divide evenly, R warns you but still returns an answer.

Your turn: use `die` to produce the numbers **2, 4, 6, 8, 10, 12** without
typing them out.

```{r vec3, exercise=TRUE}
die <- 1:6

```

```{r vec3-solution}
die <- 1:6
die * 2
```

## 4. Functions

A **function** is a name followed by parentheses. Whatever you put inside the
parentheses is what the function works on.

```{r fun1, exercise=TRUE}
die <- 1:6
mean(die)
sum(die)
round(3.1415)
factorial(3)
```

Functions can be nested --- R works from the inside out:

```{r fun2, exercise=TRUE}
round(mean(1:6))
```

### sample()

The function that makes dice possible is `sample()`. It draws at random from a
vector.

```{r fun3, exercise=TRUE}
die <- 1:6
sample(x = die, size = 1)
```

Run it several times. You get a different number each time --- you have rolled a
die.

### Arguments

`x` and `size` are the function's **arguments**. To find out what arguments a
function takes, use `args()`:

```{r fun4, exercise=TRUE}
args(sample)
args(round)
```

`round(x, digits = 0)` shows that `digits` has a **default value** of 0. You only
need to supply an argument if you want something other than the default.

```{r fun5, exercise=TRUE}
round(3.1415)
round(3.1415, digits = 2)
```

If you name your arguments, order does not matter. If you don't name them, R
matches them by position.

```{r quiz-args}
quiz(caption = "Quick check",
  question("`round(x, digits = 0)`. Which calls round pi to 2 decimal places?",
    answer("`round(3.1415, 2)`", correct = TRUE),
    answer("`round(digits = 2, x = 3.1415)`", correct = TRUE),
    answer("`round(2, 3.1415)`", message = "This rounds the number 2 to 3 decimal places."),
    answer("`round(3.1415, digits = 2)`", correct = TRUE),
    allow_retry = TRUE
  )
)
```

## 5. Rolling two dice

Now roll a **pair** of dice. Change `size` to 2:

```{r roll1, exercise=TRUE}
die <- 1:6
sample(die, size = 2)
```

Run it a few times. Do you ever see a double --- two of the same number?

You won't. By default `sample()` draws **without replacement**: once it has taken
the 3, the 3 is gone and cannot come up again. That is not how dice work. Real
dice sample **with replacement**.

```{r roll2, exercise=TRUE}
die <- 1:6
sample(die, size = 2, replace = TRUE)
```

Run that several times until you see a double.

Now sum the two dice to get the roll:

```{r roll3, exercise=TRUE}
die <- 1:6
dice <- sample(die, size = 2, replace = TRUE)
sum(dice)
```

```{r quiz-replace}
quiz(caption = "Quick check",
  question("Why did we need `replace = TRUE`?",
    answer("Without it, the same number cannot be drawn twice --- so doubles would be impossible.", correct = TRUE),
    answer("Without it, `sample()` returns an error."),
    answer("It makes the sampling faster."),
    answer("It replaces the object `die` with a new one.", message = "It has nothing to do with the object --- it's about whether a drawn value goes back in the bag."),
    allow_retry = TRUE
  )
)
```

## 6. Write your own function

Typing three lines every time you want to roll is tedious. Wrap it in a function
of your own.

The recipe is always the same:

```
name <- function(arguments) {
  body of code
}
```

Here it is for our dice:

```{r own1, exercise=TRUE}
roll <- function() {
  die <- 1:6
  dice <- sample(die, size = 2, replace = TRUE)
  sum(dice)
}

roll()
```

Run `roll()` a few times. You have just built a tool.

Two things to notice:

- The function returns the **last line** of its body. That is why `sum(dice)`
  comes last.
- `die` and `dice` live *inside* the function. They do not appear in your
  environment.

### Give it an argument

Right now `roll()` can only roll a six-sided die. Rewrite it so the die is an
**argument** with a default of `1:6` --- then it can roll anything.

```{r own2, exercise=TRUE}
roll2 <- function(______) {
  dice <- sample(______, size = 2, replace = TRUE)
  sum(dice)
}

roll2()
roll2(die = 1:20)
```

```{r own2-hint}
# The argument goes in the parentheses, with its default after an = sign.
roll2 <- function(die = 1:6) {
  ...
}
```

```{r own2-solution}
roll2 <- function(die = 1:6) {
  dice <- sample(die, size = 2, replace = TRUE)
  sum(dice)
}

roll2()
roll2(die = 1:20)
```

## 7. Roll it 10,000 times

One roll tells you nothing. `replicate()` repeats an expression as many times as
you like and collects the results.

```{r rep1, exercise=TRUE}
roll <- function() {
  die <- 1:6
  dice <- sample(die, size = 2, replace = TRUE)
  sum(dice)
}

replicate(10, roll())
```

Now do it ten thousand times and look at the shape of the results:

```{r rep2, exercise=TRUE, exercise.lines=12}
roll <- function() {
  die <- 1:6
  dice <- sample(die, size = 2, replace = TRUE)
  sum(dice)
}

rolls <- replicate(10000, roll())
hist(rolls, breaks = 1.5:12.5, main = "10,000 rolls of a fair pair of dice",
     xlab = "Sum of two dice", col = "grey80")
```

A pyramid, peaking at 7. That is what a **fair** pair of dice looks like: there
are six ways to make a 7 and only one way to make a 12.

You can also count the outcomes directly:

```{r rep3, exercise=TRUE}
roll <- function() sum(sample(1:6, size = 2, replace = TRUE))
rolls <- replicate(10000, roll())
table(rolls)
```

## 8. Weight the dice

Time to cheat.

`sample()` has one more argument we haven't used: `prob`. It takes a vector of
probabilities, one for each value, saying how likely each is to be drawn.

A **fair** die gives every side the same chance --- 1/6 each:

```{r weight1, exercise=TRUE}
sample(1:6, size = 2, replace = TRUE, prob = c(1/6, 1/6, 1/6, 1/6, 1/6, 1/6))
```

Now load it. Make the 6 come up **half** the time, and split the rest evenly
across 1 through 5 (1/10 each). The six numbers must add up to 1.

```{r weight2, exercise=TRUE}
roll <- function() {
  die <- 1:6
  dice <- sample(die, size = 2, replace = TRUE,
                 prob = c(______))
  sum(dice)
}

roll()
```

```{r weight2-hint}
# Five sides at 1/10, and the six at 5/10. Check they sum to 1:
sum(c(1/10, 1/10, 1/10, 1/10, 1/10, 5/10))
```

```{r weight2-solution}
roll <- function() {
  die <- 1:6
  dice <- sample(die, size = 2, replace = TRUE,
                 prob = c(1/10, 1/10, 1/10, 1/10, 1/10, 5/10))
  sum(dice)
}

roll()
```

### Does it work?

Ten thousand rolls of the loaded dice. Compare this histogram with the fair one
from the last section.

```{r weight3, exercise=TRUE, exercise.lines=14}
roll <- function() {
  die <- 1:6
  dice <- sample(die, size = 2, replace = TRUE,
                 prob = c(1/10, 1/10, 1/10, 1/10, 1/10, 5/10))
  sum(dice)
}

rolls <- replicate(10000, roll())
hist(rolls, breaks = 1.5:12.5, main = "10,000 rolls of weighted dice",
     xlab = "Sum of two dice", col = "indianred")
mean(rolls)
```

The peak has slid to the right and 12 is now common. The mean roll is well above
7. If you played this at a casino table, the casino would notice.

```{r quiz-weight}
quiz(caption = "Quick check",
  question("Your weighted die uses `prob = c(1/10, 1/10, 1/10, 1/10, 1/10, 5/10)`. Which sum is now the *most* likely?",
    answer("12", correct = TRUE, message = "Yes --- 12 requires two sixes, and sixes are now the most common face by far."),
    answer("7", message = "7 wins for *fair* dice. Try running the simulation and looking at `table(rolls)`."),
    answer("2"),
    answer("Every sum is equally likely."),
    allow_retry = TRUE
  )
)
```

## 9. Your turn

No hints on these. Use the console below --- everything you need is in the
sections above.

**Challenge 1.** Write `roll_n()`, a function that rolls `n` fair dice (default
2) and returns their sum.

```{r ch1, exercise=TRUE, exercise.lines=10}

```

```{r ch1-solution}
roll_n <- function(n = 2) {
  sum(sample(1:6, size = n, replace = TRUE))
}
roll_n()
roll_n(5)
```

**Challenge 2.** Weight a die so that **1 never comes up at all**, and the other
five faces are equally likely. Roll it 10,000 times and check with `table()`
that no 1s appear.

```{r ch2, exercise=TRUE, exercise.lines=10}

```

```{r ch2-solution}
weird <- replicate(10000, sample(1:6, size = 1, replace = TRUE,
                                 prob = c(0, 0.2, 0.2, 0.2, 0.2, 0.2)))
table(weird)
```

**Challenge 3.** The weighted dice in Section 8 average about 9. Can you push
the average roll all the way to **10**? Change the `prob` values, run 10,000
rolls, check the `mean()`, and try again. (How high can the average possibly go?)

```{r ch3, exercise=TRUE, exercise.lines=14}

```

## What you learned

In under an hour you used:

| Idea | Looks like |
|---|---|
| Arithmetic | `1 + 1`, `2^10`, `17 %% 5` |
| Objects | `die <- 1:6` |
| Vectors | `1:6`, element-wise math, recycling |
| Functions | `sum()`, `mean()`, `round()`, `sample()` |
| Arguments & defaults | `args(round)`, `round(x, digits = 2)` |
| Writing functions | `roll <- function() { ... }` |
| Repetition | `replicate(10000, roll())` |
| Summarising | `table()`, `hist()`, `mean()` |

That is the core of the language. Everything else is more functions.

**Where this came from:** Garrett Grolemund, *Hands-On Programming with R*,
Chapters 1--3 --- free at <https://rstudio-education.github.io/hopr/basics.html>
